Thermodynamic constraints for magic states in quantum computation

### 2021-04-15 magic_paper.tex
Intro almost done. Needs references added.

\title{Thermodynamic structures for magic states in quantum computation}

@@

\section{Introduction and background}
\label{sec:intro_prev}

\subsection{Introduction}
\label{sec:intro}

\ddd{[Still work in progress. Gotta fix narrative line first.]}

The state-injection model for quantum computation~\cite{cit:bravyi} is a central approach to realising a full-scale, fault-tolerant quantum computer. Within this scheme, Clifford unitaries can be implemented in a robust, fault-tolerant way. However, due to the Eastin-Knill theorem~\cite{Eastin_2009}, it is impossible to have a universal set of transversal gates, and although Clifford unitaries can be realised transversally, one needs to find ways around the Eastin-Knill restriction. The state-injection approach supplements the Clifford unitaries with noisy quantum states, called `magic states', that are outside the set of stabliser states and, when incorporated into Clifford circuits, provide universality. The obstacle to this is that the states are invariably noisy and so protocols must be employed to purify many copies of the magic states and improve the overall performance of the induced quantum gates. A central question is then: given a number $n$ of noisy magic states how many purified magic states can we obtain within the fault-tolerant set of quantum operations?

To address such questions, concrete distillation protocols have been developed, and there has also been analysis of bounds on distillation rates within natural theoretical frameworks. These frameworks provide theories of magic in which one views magic states as ``resource'' states with respect to a natural class of quantum operations that are consider cheap, or ``free''~\cite{Gour_2019}. One very natural class of free operations considered are those obtained from Clifford operations, measurements and the ability to discard quantum systems. However, there are several other candidate frameworks~\cite{cit:ahmadi, cit:seddon, Wang_2019}.

Magic states are necessary for achieving universal quantum computation within fault-tolerant schemes~\cite{cit:veitch, cit:mari, cit:gottesman, cit:bravyi, cit:knill, Campbell_2011}.
Identifying magic as a resource for quantum universality has led to several theories which try to provide a framework for its quantification and manipulation \cite{cit:veitch2, cit:howard, Wang_2019, Seddon_2021, Okay_2021}.
The main question that such a theory attempts to answer is:
\begin{center}
    \emph{How do physical constraints, such as hardware limitations, noise bias or thermodynamics, restrict magic distillation protocols?}
\end{center}

Such constraints make stabilizer states inequivalent, and can give rise to stabilizer states that are invariant under the constraints. In this work we shall approach this question using tools from statistical mechanics and thermodynamics. In particular, we shall view \emph{magic as a form of free energy}, relative to the set of stabilizer states. This perspective takes the convex set of stabilizer states as a manifold of ``equilibrium'' states. However, while a free energy difference exists between two thermodynamic equilibrium states at temperatures $T_1$ and $T_2$, here we focus in on the degree to which the magic state lies off the set of stabilizer states, and therefore magic is more accurately viewed as a form of non-equilibrium free energy that is strictly non-classical.

Our approach relies on the discrete Wigner representation of quantum states, in which all states and operations can be described on a discrete phase space. Crucially, we focus on magic states with negativity in their Wigner function, which is known to be a necessary condition for universality in the state-injection model. However, following a thermodynamic perspective in this context raises a novel problem, namely it is impossible to define entropies for quasi-distributions. We circumvent this obstacle by making use of a more fundamental tool in thermodynamics -- majorization theory.

Our construction also provides distillation bounds without relying on magic monotones -- instead of projecting the pre-order of magic states onto the real line (which is a monotone) we instead cover the pre-order of states with a collection of simpler `fragments', which admit analysis through the lens of majorization. The choice of fragment breaks the equivalence of all stabilizer states in the theory and can be used to consider adding physical constraints in the form of accessible protocols. \nick{We need to give some concrete example here, we keep repeating this ``advantage'' abstractly}

We provide a range of upper bounds, which reveal that the thermodynamic structure in the presence of negativity is surprisingly non-trivial and displays a range of features that do not appear in classical statistical mechanics. We also discuss how this approach could be used to construct explicit lower bounds on distillation. This in turn raises non-trivial questions that are also of interest in statistical mechanics. More precisely: what are the conditions for majorization that respects the symplectic structure of phase space?

\subsection{Prior text that might be useful}
Even though non-stabiliserness is a necessary resource for universality, it has been proven insufficient for magic state distillation~\cite{cit:bravyi, cit:campbell}.
In fact, all states with non-negative Wigner distributions have been proven to be efficiently classically simulable~\cite{cit:mari}, a result that serves as a generalization of the Gottesman-Knill theorem.
The Wigner distribution of a state in odd prime dimensions arises as the unique quasi-probability representation of quantum theory that identifies non-contextuality exactly with the states that are efficiently classically simulable~\cite{Delfosse_2017, cit:howard2, cit:veitch2}.
In this framework, the stabiliser states are the only pure states represented with non-negative distributions~\cite{cit:gross3}. 
However, there exist mixed states with non-negative Wigner distributions that are not mixtures of stabiliser states~\cite{cit:gross}.

The paper is structured as follows. In~\cref{sec:ps}, we formally define the Wigner representation of quantum states and outline its relevant properties. In~\cref{sec:frag}, we introduce majorisation and apply it on general magic theories. We go on to prove magic distillation bounds which account for abstract thermodynamic considerations in~\cref{sec:unital,sec:stab} and conclude in~\cref{sec:outlook}.

@@

The state-injection model for quantum computation~\cite{cit:bravyi} is a key avenue to realising a full-scale, fault-tolerant quantum computer. Within this approach, Clifford unitaries can be implemented in a robust, fault-tolerant way. However, due to the Eastin-Knill theorem~\cite{Eastin_2009}, it is impossible to have a universal set of transversal gates, and although Clifford unitaries can be realised transversally, one needs to find ways around the Eastin-Knill restriction. The state-injection approach supplements the Clifford unitaries with noisy quantum states, called `magic states', that are outside the set of stabliser states and, when incorporated into Clifford circuits, provide universality. The obstacle to this is that the states are invariably noisy and so protocols must be employed to purify many copies of the magic states and improve the overall performance of the induced quantum gates. A central question is then: given a number $n$ of noisy magic states how many purified magic states can we obtain within the fault-tolerant set of quantum operations?

To address such questions, concrete distillation protocols have been developed, and there has also been analysis of bounds on distillation rates within natural theoretical frameworks. These frameworks provide theories of magic in which one views magic states as ``resource'' states with respect to a natural class of quantum operations that are consider cheap, or ``free''~\cite{Gour_2019}. One very natural class of free operations considered are those obtained from Clifford operations, measurements and the ability to discard quantum systems. However, there are several other candidate frameworks~\cite{cit:ahmadi, cit:seddon, Wang_2019}.

@@

One of the most fundamental and commonly used magic monotones is the \emph{mana} of a state~\cite{cit:veitch2}, defined as
\begin{equation}
    \mana{\rho} \coloneqq \ln{(2\hspace{1pt}\sn{\rho}+1)},
\end{equation}
where the \emph{sum-negativity}~\cite{cit:veitch2} is the sum of the negative components in $\W{\rho}$,
\begin{equation}
    \sn{\rho} \coloneqq \sum\limits_{\bmx: \W[\bmx]{\rho} < 0} \abs{\W[\bmx]{\rho}}.
\end{equation}

@@

A range of monotones have been developed in the literature~\cite{cit:howard, Wang_2020, Seddon_2021} and each provides a means to upper bound distillation rates within a particular theory of magic. In this work we shall consider a broad framework for magic, with conditions that any magic theory should satisfy. However, in contrast to prior works we shall not construct magic monotones, but instead develop a generalisation to resource monotones. The approach we take will also allow us to address how distillation rates are constrained by physical properties -- such as a system having biased noise, some equilibrium structure or operations that manifest invariances, such as axial symmetry around the qubit $Z$-axis.

@@

	The location of the maximum ($x=x_{i_\star}$) varies from fragment to fragment \nick{try to calculate for thermal fragments}, but its value is independent of $\sigma$,
	\begin{align}
	L_\star &:=	\lc{\rho}{\sigma}(x_{i_\star}) 
		= \sum\limits_{\bmx: \W[\bmx]{\rho} \geq 0} \W[\bmx]{\rho} \nonumber \\
		&= 1 + \sn{\rho}.
	\end{align}

>>>>

\title{Thermodynamic constraints for magic states in quantum computation}

@@

\section{Introduction}
\label{sec:intro}

The past few years have seen rapid progress towards the goal of a fault-tolerant quantum computer\ddd{[CITE]}. However many challenges remain and there is increasing need for theory to take into account physical limitations of the hardware involved. The surface code~\cite{cit:bravyi}\ddd{[CITE MORE]} is a leading framework for fault-tolerance with very high error thresholds. model for quantum computation is a central approach to realising a full-scale, fault-tolerant quantum computer. Within this scheme, Clifford unitaries can be implemented in a robust, fault-tolerant way. However, due to the Eastin-Knill theorem~\cite{Eastin_2009}, we also know that it is impossible to have a universal set of transversal gates, and although Clifford unitaries can be realised transversally\ddd{[CSS code for this]}, one needs to find ways around the Eastin-Knill restriction. This can be achieved by injecting in quantum states, called magic states, which promote the Clifford group to universal quantum computing. However, the obstacle to this is that these states are invariably noisy and so protocols involving stabilizer operations must be employed to purify many copies of the magic states and improve the overall performance of the induced quantum gates. A central question is then the overhead on purifying many copies of a magic state into less noisy forms.

To address such questions, concrete distillation protocols have been developed, such as the Bravyi-Haah protocol that provides a quadratic reduction in noise per cycle. Such distillation rates have been improved in more recent works \ddd{[CITE]}. There is also analysis of magic protocols from the perspective of magic monotones, which provide upper bounds on distillation rates and find application in the analysis of gate synthesis \ddd{[CITE Earl+Mark paper, plus other synthesis paper(s)]}. These frameworks for magic view magic states as resource states with respect to a natural class of quantum operations that are consider cheap, or free, such as stabilizer operations~\cite{Gour_2019, cit:ahmadi, cit:seddon, Wang_2019}.

Recent work towards fault-tolerance has begun to bridge the gap between abstract theory and experiment. Extensive work has been done on error mitigation \ddd{[cite]}, and the incorporation of hardware physics into the theoretical modes. For example, the XZZX code\ddd{[cite]} is a variant of the surface code that incorporates noise bias explicitly and has been shown to attain the hashing bound of random codes.

In this work we develop a framework to analyse magic state distillation protocols where explicit physical constraints are manifest. These physical constraints are encoded in the allowed fixed-point structure of the protocol, for example a minimal temperature attainable in the device, noise-biases or fixed-point structure associated with restricted gate-sets. The approach we take is most closely aligned with resource theories of magic, however our approach differs in two ways. Firstly, we obtain distillation upper bounds without the use of monotones, and secondly our distillation bounds are functions of the physical constraints and so not directly comparable to global magic monotones.

We approach this problem using tools from statistical mechanics and recent work from the overlap of thermodynamics and quantum information theory\ddd{[cite]}. In particular, we shall view \emph{magic as a form of free energy}, relative to the set of stabilizer states. This perspective takes the convex set of stabilizer states as the set of ``equilibrium'' states, and magic corresponds to a form of non-equilibrium free energy that is strictly non-classical.

Our analysis relies on the discrete Wigner representation of quantum states, in which all states and operations can be described on a discrete phase space \ddd{[cite]}. Crucially, we focus on magic states with negativity in their Wigner representation, which is known to be a necessary condition for universality in the state-injection model. Interestingly, taking a thermodynamic perspective in this context raises a problem, namely it is impossible to define entropies for quasi-distributions. We circumvent this obstacle by making use of a more fundamental tool in thermodynamics -- majorization theory \ddd{[cite]}. To our knowledge majorization of quasi-distributions has not been considered in quantum physics before, and therefore our analysis is of broader interest to the study of non-classicality in quantum systems.

Our construction provides distillation bounds without relying on magic monotones. Instead of projecting the pre-order of magic states onto the real line (which is what a monotone involves) we instead cover the pre-order of states with a collection of simpler `fragment' pre-orders that admit analysis in terms of majorization. It is this choice of fragment that breaks the equivalence of all stabilizer states in the theory and can be used to consider adding physical constraints in the form of accessible protocols. \nick{We need to give some concrete example here, we keep repeating this ``advantage'' abstractly}

Through this approach we obtain a range of upper bounds for magic distillation rates. We provide explicit bounds for magic protocols that generate unital channels, as well as bounds that depend on the temperature and Hamiltonian of the system. We find that the thermodynamics in the presence of negativity is surprisingly non-trivial and displays a range of features that do not appear in classical statistical mechanics.

Finally, we discuss how our approach could be improved, and also used to construct explicit lower bounds on distillation that take into account physical structure. Our work also raises non-trivial questions that are also of interest in statistical mechanics. Specifically: what are the majorization conditions that respect the symplectic structure of phase space?

@@

A number of frameworks exist that provide theories of magic in which one views magic states as ``resource'' states with respect to a natural class of quantum operations that are consider cheap, or ``free''~\cite{Gour_2019}. One very natural class of free operations considered are those obtained from Clifford operations, measurements and the ability to discard quantum systems. However, there are other candidates~\cite{cit:ahmadi, cit:seddon, Wang_2019}.

@@

One of the most prominent magic monotones is the \emph{mana} of a state~\cite{cit:veitch2}, defined as
\begin{equation}
    \mana{\rho} \coloneqq \ln{(2\hspace{1pt}\sn{\rho}+1)},
\end{equation}
where the \emph{sum-negativity}~\cite{cit:veitch2} is the sum of the negative components in $\W{\rho}$,
\begin{equation}
    \sn{\rho} \coloneqq \sum\limits_{\bmx: \W[\bmx]{\rho} < 0} \abs{\W[\bmx]{\rho}}.
\end{equation}

@@

A range of monotones have been developed in the literature~\cite{cit:howard, Wang_2020, Seddon_2021} and each provides a means to upper bound distillation rates within a particular theory of magic. In this work we shall develop bounds that apply to essentially any reasonable magic theory. The central idea is to apply majorization theory to the quasi-probability distributions of magic states. In the next section we explain how these majorization constraints relate to the free operations allowed in a theory of magic.

@@

	The location of the maximum ($x=x_{i_\star}$) varies from fragment to fragment, but its value is independent of $\sigma$,
	\begin{align}
	L_\star &:=	\lc{\rho}{\sigma}(x_{i_\star}) 
		= \sum\limits_{\bmx: \W[\bmx]{\rho} \geq 0} \W[\bmx]{\rho} \nonumber \\
		&= 1 + \sn{\rho}.
	\end{align}
